---
title: "GIS Further Training in R Session 2"
author: "Julian Matthewman"
---
These "Further Training" notebooks cover the GIS further training sessions taught by Chris Grundy. If you are not familiar with the basics of GIS in R yet, check out the R notebook "Introduction to GIS". We rely once again primarily on the sf and tidyverse packages and the datasets found in the LSHTM U drive under downloads/teach/GIS/ArcGIS.

Let's start by loading our packages (installing them beforehand if necessary) and setting our working directory. Make sure this file is in the same directory the folders containing our datasets.

```{r message=FALSE, warning=FALSE}
#install.packages("sf")
#install.packages("readxl")
#install.packages("tidyverse")

library(sf)
library(readxl)
library(tidyverse)

#We use the package rstudiapi to automatically set the working directory to where this file is located.
library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Creating new variables
Here we get familiar with a few functions to create new variables in R.
Let's load the wards.xls dataset.
```{r}
ward <- read_excel("GLA-Census/wards.xls")
```
As we can see (if we view() the object) it only has 2 columns, Ward and Dha (District code).
We want to create the following 4 variables:
1. county code: this equals 01 for every ward 
2. district code: equal to the first 2 characters of the ward code
3. ons ward code: equal to the county code followed by the district code, i.e. 01ABAS
4. rha code: equal to the first character of the dha code

One of the handiest functions to manipulate tables is the dplyr (part of the tidyverse) function mutate(). Let's try and do all of this using mutate().
```{r}
ward <- ward %>% mutate(country = "01") #put the 01 in quotation marks("") so it creates strings rather than a numberical data, or else we will loose the 0 before the 1.

ward <- ward %>% mutate(district = str_sub(Ward, start = 1, end = 2)) #Here we can use the strigr (part of the tidyverse) function str_sub. We specify the variable we want to extract substrings from and the start and end characters.

ward <- ward %>% mutate(onsward = str_c(ward$country, ward$Ward)) #We use another command from the str family, this time str_c (c is for concatenate). It works like c() but for strings.

ward <- ward %>% mutate(rha = str_sub(Dha, start = 1, end = 1)) #Once again we use str_sub.
```

If we wish to export our updated data back to Excel we can do it like this:
```{r}
write_excel_csv(ward, "GLA-Census/wards_new.csv") #This will give us a .csv file, but it's easy to work with or change to .xls in Excel.
```

## Exercise
Add some more colums to the table!
```{r}
#ID number (1,2,3,4,5,...)

#Ward code in lower case letters (aaaa, aaab, aaac, ...)

#Number of "A"s in each Ward code

#Identify if Ward code contains the letter "A" (TRUE or FALSE)

```
Hint: try using seq.int(nrow(ward)), str_to_lower, str_count, str_detect

